Towards atomistic resolution structure of phosphatidylcholine glycerol backbone and choline headgroup at different ambient conditions

### 2015-02-16 HGmodelMANUSCRIPT/HGmodel_draft3.tex
updating information from blog to the manuscript
updates to HGmodel draft

\author{Alexandru Botan}
\thanks{The authors are listed in alphaphetical order.}
\thanks{The author list is not completed.}
\thanks{Institut Lumi\`ere Mati\`ere, UMR5306 Universit\'e Lyon 1-CNRS, Universit\'e de Lyon 69622 Villeurbanne, France}
%
\author{Andrea Catte}
\thanks{University of East Anglia, Norwich, United Kingdom}
\author{Fernando Favela}
\author{Patrick Fuchs}
\thanks{Institut Jacques Monod, CNRS, Universit\'e Paris Diderot, Sorbonne Paris Cit\'e, Paris, France}
%\thanks{Université Paris Diderot, Sorbonne Paris Cité, Paris, France}
\author{Matti Javanainen}
\thanks{Tampere University of Technology, Tampere, Finland}
\author{Waldemar Kulig}
\thanks{Tampere University of Technology, Tampere, Finland}
\author{Antti Lamberg}
\thanks{Department of Chemical Engineering, Kyoto University, Kyoto, Japan}
%\author{Alexander Lyubartsev}
\author{Markus S. Miettinen}
\thanks{Fachbereich Physik, Freie Universit\"at Berlin, Berlin, Germany}
\author{Luca Monticelli}
\thanks{5IBCP, CNRS UMR 5086, Lyon, France}
\author{Jukka M\"a\"att\"a}
\thanks{Aalto University, Espoo, Finland}
%\setcounter{page}{1}
\author{Vasily S. Oganesyan}
\thanks{University of East Anglia, Norwich, United Kingdom}
\author{O. H. Samuli Ollila} 
\thanks{{\bf Author to whom correspondence may be addressed. E-mail: samuli.ollila@aalto.fi.}}
\thanks{Helsinki Biophysics and Biomembrane Group, Department of Biomedical Engineering and Computational Science, Aalto University, Espoo, Finland}
\author{Marius Retegan}
\thanks{Max Planck Institute for Chemical Energy Conversion, M\"ulheim an der Ruhr, Germany}
\author{Hubert Santuz}
\thanks{INSERM, U1134, DSIMB; Institut National de la Transfusion Sanguine (INTS); Laboratoire d'Excellence GR-Ex, Paris, France}
\thanks{Universit\'e Paris Diderot, Sorbonne Paris Cit\'e, Paris, France}
\author{Joona Tynkkynen}
\thanks{Tampere University of Technology, Tampere, Finland}
%\author{Mark Wilson}
%\author{Alexander Vogel}

\begin{abstract}
Phospholipids are essential building blocks of biological membranes.
Despite of vast amount of accurate experimental data the atomistic resolution structures sampled by the glycerol backbone and choline headgroup
in phoshatidylcholine bilayers are not known. Atomistic resolution molecular dynamics simulation model 
would automatically resolve the structures giving an interpretation of experimental results, if the model
reproduced the experimental data. In this work we compare the C--H bond vector order
parameters for glycerol backbone and choline headgroup between 14 different atomistic resolution
models and experiments in a fully hydrated lipid bilayer. The current models are not accurately enough to resolve the structure.
However, closer inspection of three best performing models (CHARMM36, GAFFlipid and MacRog) suggest
that improvements in the sampled dihedral angle distributions would potentilly lead to the model which
would resolve the structure. Despite of the inaccuracy in the fully hydrated
structures, the response to dehydration, i.e. the tilting of the P--N vector more parallel to membrane normal, 
is qualitatively correct in all models. The CHARMM36 and MacRog models describe the interactions
between lipids and cholesterol better than the Berger/H\"oltje model.
This work has been, and continues to be, progressed and discussed through the blog: \url{nmrlipids.blogspot.fi}. 
Everyone is invited to join the discussion and make contributions through the blog. 
The manuscript will be eventually submitted to an appropriate scientific journal. 
Everyone who has contributed to the work through the blog will be offered 
coauthorship. For more details see: \url{nmrlipids.blogspot.fi}.

@@

In this work we first review the most relevant experimental data for the glycerol backbone and choline headgroup order parameters
in a phosphatidylcholine lipid bilayer. Then the available atomistic resolution lipid models are carefully compared to the 
experimental data. The comparison reveals that the CHARMM36~\cite{klauda10}, GAFFlipid~\cite{dickson12} and MacRog models~\cite{maciejewski14}
have the most realistic glycerol backbone and choline structures. We also compare the glycerol backbone and choline 
structures between the most often used (Berger based) lipid model~\cite{berger97} 
\todo{Based on the results and discussions with Olle Edholm and Peter Tieleman in the blog, it seems that 
the glycerol backbone dihedrals are not the same in the model used in the publication by Berger et al.~\cite{berger97}
and in the model shared in the webpage of Peter Tieleman (http://wcm.ucalgary.ca/tieleman/downloads).
In this work (as it is quite often done in other works as well) we currently refer to the model in the webpage as ''Berger'' model.
This is misleading. We have to call it something else. One suggestion could be to call it ''BergerM'', because it was firt time
used in Marrink et al. Biophys. J. 74,931-943 (1998). Any other suggestions?\\
COMMENT: Patrick Fuchs suggested ''Berger-T''. This is better. \\
COMMENT2: Current trial, based in blog discussion, is to use Berger-MOLECULE-YEAR [citation].
}
and the best performing models, to demonstrate that by using the 
order parameters we can distinguish the more reasonable structures from the less reasonable ones. However, none of the current models 
is accurate enough to properly resolve the atomistic resolution stuctures.

@@

\subsection{Analysis}
\todo{ We should mention the sign issue and refer to the blog post:
http://nmrlipids.blogspot.fi/2014/04/on-signs-of-order-parameters.html somewhere.
Maybe we should change the name of this section II.B. to: "Analysis and terminology of order parameters from experiments 
and simulations" or similar, and also move the "forking" explanation here.}
The order parameter of a hydrocarbon C--H vector is defined as $S_{{\rm CH}}=\frac{3}{2}\langle \cos^2 \theta-1 \rangle$, where
the angle brackets denote an ensemble average over the sampled conformations, and $\theta$ is the angle between the C--H bond and the membrane normal.
The order parameters can be measured by detecting quadrupolar splitting with $^2$H NMR~\cite{seelig77c} or by detecting dipolar 
splitting with $^{13}$C NMR~\cite{hong95a,gross97,dvinskikh05a,ferreira13}. The measurements are based on
different physical interactions and also the connection between order parameters and quadrupolar or dipolar splitting
are different. The order parameters from the measured quadrupolar splitting $\Delta \nu_Q$ ($^2$H NMR) are calculated using 
the equation $|S_{{\rm CD}}|=\frac{4}{3} \frac{e^2qQ}{h} \Delta \nu_Q$, where the value for the static quadrupole
splitting constant is estimated from various experiments to be 170~kHz leading to a numerical relation $|S_{{\rm CD}}|=0.00784 \times \Delta \nu_Q$~\cite{seelig77c}. 
The order parameters from the measured dipolar splitting $d_\mathrm{CH}$ ($^{13}$C NMR) are calculated using equation
$|S_{{\rm CH}}|=\frac{4\pi\langle r_\mathrm{CH}^3 \rangle}{\hbar \mu_0 \gamma_h \gamma_c} d_\mathrm{CH}$, where
values between 20.2--22.7 kHz are used for $\frac{4\pi\langle r_\mathrm{CH}^3 \rangle}{\hbar \mu_0 \gamma_h \gamma_c}$,
depending on the original authors~\cite{hong95a,gross97,dvinskikh05a,ferreira13}.
It is important to note that the order parameters measured with different techniques based on different physical interactions are in good agreement
with each other (see Results and Discussion), indicating very high quantitative accuracy of the measurements.
For a more detailed discussion see \url{http://nmrlipids.blogspot.fi/2014/02/accuracy-of-order-parameter-measurements.html}
\todo{This is temporarily linked to the blog. Currently I think that the posts (with comments included) to which we want to cite are put into 
the Figshare in pdf format as they are in when the publication is submitted. Figshare allows commenting to continue the discussion.}.

The order parameters from simulations were calculated directly using the definition.
%$S_{{\rm CH}}=\frac{3}{2} \langle \cos^2 \theta-1 \rangle$. 
For the united atom models the hydrogen locations were generated 
in the trajectories post-simulationally using the positions of the heavy atoms and the known hydrocarbon geometries.
The statistical errors were estimated for the best performing simulation
models by calculating the error of the mean for the average over individual lipids in the system.

@@

\begin{table*}[htb]

@@

\caption{Simulated single component fully hydrated lipid bilayers. The simulation file data sets marked with $^*$ also include a part of the trajectory.

@@

  are given in the Supplementary Information. 

@@

CHARMM36\cite{klauda10}       & POPC   & 72  &  2242 & 303 & 30 & 20  & \cite{charmm36filesSHORT}$^*$ & SI \\
CHARMM36\cite{klauda10}      & POPC   & 128 &  5120    & 303 & 150 & 100  & \cite{charmm36files}$^*$ \todoi{DONE}   & SI \\
CHARMM36\cite{klauda10}       & DPPC   & 72  &  2189 & 323 & 30 & 25  & ?\todoi{Samuli, put to Zenodo}  & SI \\

@@

MacRog\cite{maciejewski14}  & POPC & 288  & 12600 & 310 & 100 & 80  & \cite{macrogFILES}$^*$ & SI  \\
GAFFlipid\cite{dickson12}       & POPC & 126  & 3948  & 303 & 137 & 32  & \cite{GAFFlipidFILES}$^*$ & SI \\
GAFFlipid\cite{dickson12}       & DPPC & 72  & 2197  & 323 & 90 & 50  & \cite{GAFFlipidFILESdppc}$^*$ & SI \\
Lipid14\cite{dickson14}         & POPC  & 72 & 2234 & 303 & 100 & 50  & \cite{lipid14files}$^*$ & SI \\
Poger\cite{poger10}             & DPPC  & 128 & 5841 & 323 & 2$\times$100 & 2$\times$50 & \cite{pogerFILESpme1,pogerFILESpme2}$^*$\todoi{DONE} & SI \\
Slipid\cite{jambeck12}          & DPPC & 128 & 3840 & 323 & 150 & 100 & \cite{slipidsFILES}$^*$ & SI \\
Slipid\cite{jambeck12}          & POPC & 128 & 5120 & 303 & 200 & 150 & \cite{slipidsFILESpopc}$^*$ & SI \\
Kukol\cite{kukol09}          & POPC   & 512 & 20564 & 298 & 50 & 30  & \cite{kukolFILES}$^*$ & SI \\
Chiu et al.\cite{chiu09} \todoi{Samuli, put to Zenodo}     & POPC  & 128 & 3552  & 298 & 56 & 50  & ? & SI \\
H\"ogberg08\cite{rabinovich14}\todoi{Alexander Luybartsev, let us know if you share the files.}  & POPC   &  128 & 3840  & 303 & 100 & 80  & ? & \cite{rabinovich14}  \\
H\"ogberg08\cite{hogberg08}\todoi{Alexander Luybartsev, let us know if you share the files.}  & DMPC   &  98 & 3840  & 303 & 75 & 50 & ? & \cite{hogberg08} \\
Ulmschneiders\cite{Ulmschneider09}    & POPC  & 128 & 3328 & 310 & 100 & 50 & \cite{ulmschneiderFILES}$^*$ & SI \\
Tj\"ornhammar et al.\cite{tjornhammar14}   & DPPC  & 144 & 7056 & 323 & 200 & 100 & \cite{tjornhammarfiles}$^*$ & \cite{tjornhammar14} \\

@@

\subsection{Full hydration: Experimental order parameters for the glycerol backbone and headgroup}\label{experiments}
The specific deuteration of $\alpha$-, $\beta$- and g$_3$- segments of dipalmitoylphosphatidylcholine (DPPC) has been successful, 
allowing the order parameter measurements for these segments by $^2$H--NMR~\cite{gally75,brown77,brown78,akutsu81}.
In addition, the order parameters for all glycerol backbone and choline headgroup segments in egg yolk lecithin~\cite{hong95a},
1,2-dimyristoyl-sn-glycero-3-phosphocholine (DMPC)~\cite{hong95b,gross97,dvinskikh05a}, 
1,2-dioleoyl-sn-glycero-3-phosphocholine (DOPC)~\cite{warschawski05} and POPC~\cite{warschawski05,ferreira13}
have been measured with several different implementations of $^{13}$C NMR experiments.
The experimental absolute values of glycerol backbone and choline order parameters from various publications are shown in Fig.~\ref{HGorderparameters}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{../DATAreportediINblog/comparisonSorted.pdf}
\todo{Markus Miettinen suggested that we should consider making one more figure where only experimental data would be shown and that would be discussed in Section~\ref{experiments}} \\
\todo{I think that this figure is now very good. There is also the interactive version by Hubert Santuz now in https://plot.ly/~HubertSantuz/72/lipid-force-field-comparison/
we should figure out which is the most practical way to put that behind permalink once it finalized (Zenodo, figshare or something else?) and then put a citation in the paper.}
  \caption{\label{HGorderparameters}
  Order parameteres from simulations listed in Table~\ref{systems} and experiments for glycerol and choline groups.
  The experimental values were taken from the following publications: DMPC 303~K from \cite{gross97}, DMPC 314~K from \cite{dvinskikh05a}, DPPC 310K from \cite{gally75}, 
  DPPC 323~K from \cite{akutsu81}, and POPC 298~K \cite{ferreira13}.
  The vertical bars shown for some of the computational values are not error bars, but demonstrate that for 
  these systems we had two data sets; the ends of the bars mark the values from the two sets, and the dot marks their measurement-time-weighted average. 
  The H{\"o}gberg et al. force field in Table~\ref{systems} is abbreviated in this figure as HoNiLy, Tj{\"o}rnhammar et al. as TjoEdh, CHARMM36-UA as C36UA,
  Rabinovic et al. as RabLyu and Ulmschneider and Ulmschneider as UlmUlm.
} 
\end{figure}

@@

In addition to the numerical values, an important feature of the glycerol backbone is the 
inequality of order parameters for the two hydrogens attached to the same carbon in g$_1$ and g$_3$ segments,
while the hydrogens in choline $\alpha$ and $\beta$ segments give equal values.
Note that in this work we call the phenomena of unequal order parameters for hydrogens attached to the same carbon as ``forking'' to avoid 
confusion with dipolar and quadrupolar splitting in NMR terminology. Forking is also observed experimentally for the C$_2$ carbon in the \textit{sn}-2
chain of all phosholipids, and it is known to arise from differently sampled orientations of the two C--H bonds, not from two 
separate populations of lipid conformations~\cite{engel81}. The forking in glycerol backbone g$_3$ segment is small ($\approx$ 0.02) 
and some experiments only report the larger value or the average value~\cite{akutsu81,ferreira13}. 
In contrast, forking is significant for the glycerol backbone g$_1$ segment, whose lower order parameter is close to zero and the
larger one has an absolute value of approximately 0.13--0.15. Forking was studied in detail by Gally et al.~\cite{gally81}, who used E. Coli to 
stereospecifically deuterate the different hydrogens attached to the g$_1$ or g$_3$ groups in PE lipids, and measured the order parameters from the lipid 
extract. This experiment gave the lower order parameter when deuterium was in the S position of g$_1$ or R position for g$_3$.
Since the glycerol backbone order parameters are very similar irrespective of the headgroup chemistry (PC,PE and PG) or lipid 
environment~\cite{gally81}, it is reasonable to assume that the stereospecifity measured for the PE lipids
holds also for the PC lipids.

In Fig.~\ref{HGorderparameters} we have shown the absolute values of order parameters as these are accessible
with both $^2$H NMR and $^{13}$C NMR techniques. However, $^{13}$C NMR techniques allow also the measurement of 
the sign of the order parameter~\cite{hong95a,hong95b,gross97}. The measured sign is negative for almost all the carbons 
discussed in this work as only $\alpha$ is positive~\cite{hong95a,hong95b,gross97}.

>>>>

\author{Alexandru Botan}
\thanks{The authors are listed in alphaphetical order.}
\thanks{The author list is not completed.}
\thanks{Institut Lumi\`ere Mati\`ere, UMR5306 Universit\'e Lyon 1-CNRS, Universit\'e de Lyon 69622 Villeurbanne, France}
%
%\author{Andrea Catte}
%\thanks{University of East Anglia, Norwich, United Kingdom}
\author{Fernando Favela}
\author{Patrick Fuchs}
\thanks{Institut Jacques Monod, CNRS, Universit\'e Paris Diderot, Sorbonne Paris Cit\'e, Paris, France}
%\thanks{Université Paris Diderot, Sorbonne Paris Cité, Paris, France}
\author{Matti Javanainen}
\thanks{Tampere University of Technology, Tampere, Finland}
\author{Waldemar Kulig}
\thanks{Tampere University of Technology, Tampere, Finland}
\author{Antti Lamberg}
\thanks{Department of Chemical Engineering, Kyoto University, Kyoto, Japan}
%\author{Alexander Lyubartsev}
\author{Markus S. Miettinen}
\thanks{Fachbereich Physik, Freie Universit\"at Berlin, Berlin, Germany}
\author{Luca Monticelli}
\thanks{5IBCP, CNRS UMR 5086, Lyon, France}
\author{Jukka M\"a\"att\"a}
\thanks{Aalto University, Espoo, Finland}
%\setcounter{page}{1}
%\author{Vasily S. Oganesyan}
%\thanks{University of East Anglia, Norwich, United Kingdom}
\author{O. H. Samuli Ollila} 
\thanks{{\bf Author to whom correspondence may be addressed. E-mail: samuli.ollila@aalto.fi.}}
\thanks{Helsinki Biophysics and Biomembrane Group, Department of Biomedical Engineering and Computational Science, Aalto University, Espoo, Finland}
\author{Marius Retegan}
\thanks{Max Planck Institute for Chemical Energy Conversion, M\"ulheim an der Ruhr, Germany}
\author{Tomasz Rog}
\thanks{Tampere University of Technology, Tampere, Finland}
\author{Hubert Santuz}
\thanks{INSERM, U1134, DSIMB; Institut National de la Transfusion Sanguine (INTS); Laboratoire d'Excellence GR-Ex, Paris, France}
\thanks{Universit\'e Paris Diderot, Sorbonne Paris Cit\'e, Paris, France}
\author{Joona Tynkkynen}
\thanks{Tampere University of Technology, Tampere, Finland}
%\author{Mark Wilson}
%\author{Alexander Vogel}

\begin{abstract}
Phospholipids are essential building blocks of biological membranes.
Despite of vast amount of accurate experimental data, the atomistic resolution structures sampled by the glycerol backbone and choline headgroup
in phoshatidylcholine bilayers are not known. Atomistic resolution molecular dynamics simulations model 
would automatically resolve the structures giving an interpretation of experimental results, if the model
reproduced the experimental data. In the present work we simulate phosphatidylcholine (PC) lipid bilayers with 13 different
atomistic models, and we compare simulations with experiments (in fully hydrated conditions) in terms of C--H bond vector order
parameters for the glycerol backbone and choline headgroups. None of the current models is not accurately enough to resolve the structure with experimental accuracy.
However, closer inspection of three best performing models (CHARMM36, GAFFlipid and MacRog) suggest
that improvements in the sampled dihedral angle distributions would potentilly lead to the model which
would resolve the structure. Despite of the inaccuracy in the fully hydrated
structures, the response to dehydration, i.e. the tilting of the P--N vector more parallel to membrane normal, 
is qualitatively correct in all models. The CHARMM36 and MacRog models describe the interactions
between lipids and cholesterol better than the Berger/H\"oltje model.
This work has been, and continues to be, progressed and discussed through the blog: \url{nmrlipids.blogspot.fi}. 
Everyone is invited to join the discussion and make contributions through the blog. 
The manuscript will be eventually submitted to an appropriate scientific journal. 
Everyone who has contributed to the work through the blog will be offered 
coauthorship. For more details see: \url{nmrlipids.blogspot.fi}.

@@

In this work we first review the most relevant experimental data for the glycerol backbone and choline headgroup order parameters
in a phosphatidylcholine lipid bilayer. Then the available atomistic resolution lipid models are carefully compared to the 
experimental data. The comparison reveals that the CHARMM36~\cite{klauda10}, GAFFlipid~\cite{dickson12} and MacRog models~\cite{maciejewski14}
have the most realistic glycerol backbone and choline structures. We also compare the glycerol backbone and choline 
structures between the most often used (Berger based) lipid model~\cite{berger97} 
\todo{If there are no objections, we will leave the naming convention of different Berger versions as it is now.
%Based on the results and discussions with Olle Edholm and Peter Tieleman in the blog, it seems that 
%the glycerol backbone dihedrals are not the same in the model used in the publication by Berger et al.~\cite{berger97}
%and in the model shared in the webpage of Peter Tieleman (http://wcm.ucalgary.ca/tieleman/downloads).
%In this work (as it is quite often done in other works as well) we currently refer to the model in the webpage as ''Berger'' model.
%This is misleading. We have to call it something else. One suggestion could be to call it ''BergerM'', because it was firt time
%used in Marrink et al. Biophys. J. 74,931-943 (1998). Any other suggestions?\\
%COMMENT: Patrick Fuchs suggested ''Berger-T''. This is better. \\
%COMMENT2: Current trial, based in blog discussion, is to use Berger-MOLECULE-YEAR [citation].
}
and the best performing models, to demonstrate that by using the 
order parameters we can distinguish the more reasonable structures from the less reasonable ones. However, none of the current models 
is accurate enough to properly resolve the atomistic resolution stuctures.

@@

\subsection{Order parameters from experiments}
\todo{DONE
%We should mention the sign issue and refer to the blog post:
%http://nmrlipids.blogspot.fi/2014/04/on-signs-of-order-parameters.html somewhere.
%Maybe we should change the name of this section II.B. to: "Analysis and terminology of order parameters from experiments 
%and simulations" or similar, and also move the "forking" explanation here.
}
The order parameter of a hydrocarbon C--H vector is defined as 
\begin{equation}\label{orderP}
S_{{\rm CH}}=\frac{3}{2}\langle \cos^2 \theta-1 \rangle,
\end{equation} 
where the angle brackets denote an ensemble average over the sampled conformations, and $\theta$ is the angle between the C--H bond and the membrane normal.
The absolute values of order parameters can be measured by detecting quadrupolar splitting with $^2$H NMR~\cite{seelig77c} or by detecting dipolar 
splitting with $^{13}$C NMR~\cite{hong95a,gross97,dvinskikh05a,ferreira13}. The measurements are based on
different physical interactions and also the connection between order parameters and quadrupolar or dipolar splitting
are different. The absolute values of order parameters from the measured quadrupolar splitting $\Delta \nu_Q$ ($^2$H NMR) are calculated using 
the equation $|S_{{\rm CD}}|=\frac{4}{3} \frac{e^2qQ}{h} \Delta \nu_Q$, where the value for the static quadrupole
splitting constant is estimated from various experiments to be 170~kHz leading to a numerical relation $|S_{{\rm CD}}|=0.00784 \times \Delta \nu_Q$~\cite{seelig77c}. 
The absolute values of order parameters from the measured dipolar splitting $d_\mathrm{CH}$ ($^{13}$C NMR) are calculated using equation
$|S_{{\rm CH}}|=\frac{4\pi\langle r_\mathrm{CH}^3 \rangle}{\hbar \mu_0 \gamma_h \gamma_c} d_\mathrm{CH}$, where
values between 20.2--22.7 kHz are used for $\frac{4\pi\langle r_\mathrm{CH}^3 \rangle}{\hbar \mu_0 \gamma_h \gamma_c}$,
depending on the original authors~\cite{hong95a,gross97,dvinskikh05a,ferreira13}.
It is important to note that the order parameters measured with different techniques based on different physical interactions are in good agreement
with each other (see Results and Discussion), indicating very high quantitative accuracy of the measurements.
For a more detailed discussion see \url{http://nmrlipids.blogspot.fi/2014/02/accuracy-of-order-parameter-measurements.html}
\todo{This is temporarily linked to the blog. Currently I think that the posts (with comments included) to which we want to cite are put into 
the Figshare in pdf format as they are in when the publication is submitted. Figshare allows commenting to continue the discussion.
COMMENT: There is also the option to put the most relevant blog posts as a supplementary material. The advantage of this would
be that the posts would be peer reviewed as well.}.

{\bf The absolute values of order parameters are accessible with both $^2$H NMR and $^{13}$C NMR techniques. 
However, only $^{13}$C NMR techniques allow also the measurement of the sign of the order parameter~\cite{hong95a,hong95b,gross97}. 
The measured sign is negative for almost all the carbons discussed in this work as only $\alpha$ is positive~\cite{hong95a,hong95b,gross97}. 
For more detailed discussion about the sign measurements of the order parameters see \url{http://nmrlipids.blogspot.fi/2014/04/on-signs-of-order-parameters.html}

For most CH$_2$ segments in fluid phosholipid bilayer the order parameters are equal for both hydrogens attached to the same carbon.
However, in some cases (e.g. g$_1$, g$_3$ and  C$_2$ carbon in the \textit{sn}-2 chain) the order parameters are not equal and this 
can be observed with both $^2$H NMR and $^{13}$C NMR techniques. In the present work we call the phenomena of unequal order parameters 
for hydrogens attached to the same carbon as {\it forking} to avoid confusion with dipolar and quadrupolar splitting in NMR terminology. 
Forking has been studied in detail with $^2$H NMR techniques by deuterating the R or S position in CH$_2$ segment, and
the studies show that the forking arises from differently sampled orientations of the two C--H bonds, not from two 
separate populations of lipid conformations~\cite{engel81,gally81}.
}

\subsection{Order parameters from simulations}
The order parameters from simulations were calculated directly using the definition as in Eq.~\ref{orderP}.
%$S_{{\rm CH}}=\frac{3}{2} \langle \cos^2 \theta-1 \rangle$. 
For the united atom models the hydrogen positions were generated 
in the trajectories post-simulationally using the positions of the heavy atoms and the known hydrocarbon geometries.
{\bf For the statistical error estimates, the time average of order parameters were first calculated separately
for each lipid in the system. Then it was assumed that different lipids are statistically independent 
entities (which should be the case in fluid phase) and the error of the mean for the average over individual 
lipids in the system was calculated and used as error bar for the order parameters.}

@@

\begin{table*}[]

@@

\caption{Simulated single component fully hydrated lipid bilayer systems. The simulation file data sets marked with $^*$ also include a part of the trajectory.

@@

  are given in the Supplementary Information. The abbreviations are 1-palmitoyl-2-oleoylphosphatidylcholine (POPC), dipalmitoylphosphatidylcholine (DPPC), 1,2-dimyristoyl-sn-glycero-3-phosphocholine (DMPC),
1,2-dioleoyl-sn-glycero-3-phosphocholine (DOPC) and dilauroylphosphatidylcholine (DLPC).

@@

CHARMM36~\cite{klauda10}       & POPC   & 72  &  2242 & 303 & 30 & 20  & \cite{charmm36filesSHORT}$^*$ & SI \\
CHARMM36~\cite{klauda10}      & POPC   & 128 &  5120    & 303 & 150 & 100  & \cite{charmm36files}$^*$ \todoi{DONE}   & SI \\
CHARMM36~\cite{klauda10}       & DPPC   & 72  &  2189 & 323 & 30 & 25  & ?\todoi{Samuli, put to Zenodo}  & SI \\

@@

MacRog~\cite{maciejewski14}  & POPC & 288  & 12600 & 310 & 100 & 80  & \cite{macrogFILES}$^*$ & SI  \\
GAFFlipid~\cite{dickson12}       & POPC & 126  & 3948  & 303 & 137 & 32  & \cite{GAFFlipidFILES}$^*$ & SI \\
GAFFlipid~\cite{dickson12}       & DPPC & 72  & 2197  & 323 & 90 & 50  & \cite{GAFFlipidFILESdppc}$^*$ & SI \\
Lipid14 \cite{dickson14}         & POPC  & 72 & 2234 & 303 & 100 & 50  & \cite{lipid14files}$^*$ & SI \\
Poger \cite{poger10}             & DPPC  & 128 & 5841 & 323 & 2$\times$100 & 2$\times$50 & \cite{pogerFILESpme1,pogerFILESpme2}$^*$\todoi{DONE} & SI \\
Slipid \cite{jambeck12}          & DPPC & 128 & 3840 & 323 & 150 & 100 & \cite{slipidsFILES}$^*$ & SI \\
Slipid \cite{jambeck12}          & POPC & 128 & 5120 & 303 & 200 & 150 & \cite{slipidsFILESpopc}$^*$ & SI \\
Kukol \cite{kukol09}          & POPC   & 512 & 20564 & 298 & 50 & 30  & \cite{kukolFILES}$^*$ & SI \\
Chiu \cite{chiu09} \todoi{Samuli, put to Zenodo}     & POPC  & 128 & 3552  & 298 & 56 & 50  & ? & SI \\
H\"ogberg08 \cite{rabinovich14}\todoi{Alexander Luybartsev, let us know if you share the files.}  & POPC   &  128 & 3840  & 303 & 100 & 80  & ? & \cite{rabinovich14}  \\
H\"ogberg08 \cite{hogberg08}\todoi{Alexander Luybartsev, let us know if you share the files.}  & DMPC   &  98 & 3840  & 303 & 75 & 50 & ? & \cite{hogberg08} \\
Ulmschneiders \cite{Ulmschneider09}    & POPC  & 128 & 3328 & 310 & 100 & 50 & \cite{ulmschneiderFILES}$^*$ & SI \\
Tj\"ornhammar14 \cite{tjornhammar14}   & DPPC  & 144 & 7056 & 323 & 200 & 100 & \cite{tjornhammarfiles}$^*$ & \cite{tjornhammar14} \\

@@

\subsection{Full hydration: Experimental order parameters for the glycerol backbone and headgroup}\label{experiments}
The specific deuteration of $\alpha$-, $\beta$- and g$_3$- segments of DPPC has been successful, 
allowing the absolute value order parameter measurements for these segments by $^2$H--NMR~\cite{gally75,brown77,brown78,akutsu81}.
In addition, the absolute values of order parameters for all glycerol backbone and choline headgroup segments in egg yolk lecithin~\cite{hong95a},
DMPC~\cite{hong95b,gross97,dvinskikh05a}, DOPC~\cite{warschawski05} and POPC~\cite{warschawski05,ferreira13}
have been measured with several different implementations of $^{13}$C NMR experiments. In addition, the signs of order parameters in some systems
are measured with $^{13}$C NMR tehcniques~\cite{hong95a,hong95b,gross97}. The experimental absolute values of glycerol backbone 
and choline order parameters from various publications are shown in Fig.~\ref{HGorderparameters}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{../DATAreportediINblog/comparisonSorted.pdf}
\todo{Markus Miettinen suggested that we should consider making one more figure where only experimental data would be shown and that would be discussed in Section~\ref{experiments}. \\
COMMENT: I do not think that we need such figure anymore since this figure is more clear now.} \\
\todo{I think that this figure is now very good. There is also the interactive version by Hubert Santuz now in https://plot.ly/~HubertSantuz/72/lipid-force-field-comparison/
we should figure out which is the most practical way to put that behind permalink once it finalized (Zenodo, figshare or something else?) and then put a citation in the paper.}
  \caption{\label{HGorderparameters}
  Order parameteres from simulations listed in Table~\ref{systems} and experiments for glycerol and choline groups.
  The experimental values were taken from the following publications: DMPC 303~K from \cite{gross97}, DMPC 314~K from \cite{dvinskikh05a}, DPPC 310K from \cite{gally75}, 
  DPPC 323~K from \cite{akutsu81}, and POPC 298~K \cite{ferreira13}.
  The vertical bars shown for some of the computational values are not error bars, but demonstrate that for 
  these systems we had two data sets; the ends of the bars mark the values from the two sets, and the dot marks their measurement-time-weighted average. 
  The H{\"o}gberg et al. force field in Table~\ref{systems} is abbreviated in this figure as HoNiLy, Tj{\"o}rnhammar et al. as TjoEdh, CHARMM36-UA as C36UA,
  Rabinovic et al. as RabLyu and Ulmschneider and Ulmschneider as UlmUlm.
} 
\end{figure}

@@

In addition to the numerical values, an important feature of the glycerol backbone is the 
inequality of order parameters for the two hydrogens attached to the same carbon in g$_1$ and g$_3$ segments,
while the hydrogens in choline $\alpha$ and $\beta$ segments give equal values.
The forking in glycerol backbone g$_3$ segment is small ($\approx$ 0.02) 
and some experiments only report the larger value or the average value~\cite{akutsu81,ferreira13}. 
In contrast, forking is significant for the glycerol backbone g$_1$ segment, whose lower order parameter is close to zero and the
larger one has an absolute value of approximately 0.13--0.15. Forking was studied in detail by Gally et al.~\cite{gally81}, who used E. Coli to 
stereospecifically deuterate the different hydrogens attached to the g$_1$ or g$_3$ groups in PE lipids, and measured the order parameters from the lipid 
extract. This experiment gave the lower order parameter when deuterium was in the S position of g$_1$ or R position for g$_3$.
Since the glycerol backbone order parameters are very similar irrespective of the headgroup chemistry (PC,PE and PG) or lipid 
environment~\cite{gally81}, it is reasonable to assume that the stereospecifity measured for the PE lipids
holds also for the PC lipids.
